---
title: "15 Insights from Oklahoma School Enrollment Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{15 Insights from Oklahoma School Enrollment Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)
```

```{r load-packages}
library(okschooldata)
library(dplyr)
library(tidyr)
library(ggplot2)
theme_set(theme_minimal(base_size = 14))
```

Oklahoma educates over 700,000 students across 540 school districts. This vignette explores key enrollment trends and patterns using data from the Oklahoma State Department of Education.

## 1. Statewide Enrollment Trend

Oklahoma's enrollment has been growing since the COVID dip in 2021, adding roughly 25,000 students.

```{r statewide-chart}
# Fetch statewide enrollment over time
enr_state <- fetch_enr_multi(2016:2025) |>
  filter(is_state, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  select(end_year, n_students)

ggplot(enr_state, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.2, color = "#841617") +
  geom_point(size = 3, color = "#841617") +
  scale_y_continuous(labels = scales::comma, limits = c(650000, NA)) +
  scale_x_continuous(breaks = 2016:2025) +
  labs(
    title = "Oklahoma Statewide Enrollment",
    subtitle = "Public school enrollment 2016-2025",
    x = "School Year (End)",
    y = "Students"
  )
```

## 2. Top Districts Comparison

Oklahoma City and Tulsa are the largest districts, followed by suburban Edmond and Moore.

```{r top-districts-chart}
# Top 6 districts by enrollment
top_districts <- c("55I001", "72I001", "14I004", "14I002", "09I001", "31I001")
district_names <- c(
  "55I001" = "Oklahoma City",
  "72I001" = "Tulsa",
  "14I004" = "Edmond",
  "14I002" = "Moore",
  "09I001" = "Broken Arrow",
  "31I001" = "Lawton"
)

enr_top <- fetch_enr_multi(2016:2025) |>
  filter(
    district_id %in% top_districts,
    is_district,
    subgroup == "total_enrollment",
    grade_level == "TOTAL"
  ) |>
  mutate(district_label = district_names[district_id])

ggplot(enr_top, aes(x = end_year, y = n_students, color = district_label)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(breaks = seq(2016, 2025, 2)) +
  scale_color_brewer(palette = "Set2") +
  labs(
    title = "Oklahoma's Largest School Districts",
    subtitle = "Enrollment trends 2016-2025",
    x = "School Year (End)",
    y = "Students",
    color = "District"
  ) +
  theme(legend.position = "bottom")
```

## 3. Demographic Breakdown

Oklahoma's student population is diverse, with significant Native American, Hispanic, and multiracial populations.

```{r demographics-chart, eval=FALSE}
# Statewide demographics for latest year
# NOTE: Demographic breakdown data not available in OSDE enrollment files
# This chart is disabled pending data source updates
demo_subgroups <- c("white", "black", "hispanic", "asian",
                    "native_american", "pacific_islander", "multiracial")

enr_demo <- fetch_enr(2025) |>
  filter(is_state, grade_level == "TOTAL", subgroup %in% demo_subgroups) |>
  mutate(
    subgroup_label = case_when(
      subgroup == "white" ~ "White",
      subgroup == "black" ~ "Black",
      subgroup == "hispanic" ~ "Hispanic",
      subgroup == "asian" ~ "Asian",
      subgroup == "native_american" ~ "Native American",
      subgroup == "pacific_islander" ~ "Pacific Islander",
      subgroup == "multiracial" ~ "Two or More Races",
      TRUE ~ subgroup
    ),
    subgroup_label = factor(subgroup_label,
      levels = c("White", "Hispanic", "Native American", "Black",
                 "Two or More Races", "Asian", "Pacific Islander"))
  )

ggplot(enr_demo, aes(x = reorder(subgroup_label, -n_students), y = n_students, fill = subgroup_label)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = scales::percent(pct, accuracy = 0.1)), vjust = -0.5, size = 3.5) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.15))) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Oklahoma Student Demographics (2024-25)",
    subtitle = "Enrollment by race/ethnicity",
    x = NULL,
    y = "Students"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## 4. Urban vs Suburban Growth

While Oklahoma City has lost students, suburban districts like Edmond and Moore have grown steadily.

```{r growth-chart}
# Compare OKC metro area trends
metro_districts <- c("55I001", "14I004", "14I002")
metro_names <- c(
  "55I001" = "Oklahoma City",
  "14I004" = "Edmond",
  "14I002" = "Moore"
)

enr_metro <- fetch_enr_multi(2016:2025) |>
  filter(
    district_id %in% metro_districts,
    is_district,
    subgroup == "total_enrollment",
    grade_level == "TOTAL"
  ) |>
  mutate(district_label = metro_names[district_id]) |>
  group_by(district_label) |>
  mutate(
    baseline = first(n_students),
    change_pct = (n_students - baseline) / baseline * 100
  ) |>
  ungroup()

ggplot(enr_metro, aes(x = end_year, y = change_pct, color = district_label)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = seq(2016, 2025, 2)) +
  scale_color_manual(values = c("Oklahoma City" = "#841617", "Edmond" = "#1a5276", "Moore" = "#148f77")) +
  labs(
    title = "Urban vs Suburban Enrollment Growth",
    subtitle = "Percent change since 2016 in OKC metro area",
    x = "School Year (End)",
    y = "Percent Change from 2016",
    color = "District"
  ) +
  theme(legend.position = "bottom")
```

## 5. Grade Level Distribution

Kindergarten enrollment dropped sharply during COVID and is still recovering.

```{r grade-chart}
# Grade level enrollment over time
grade_levels <- c("K", "01", "05", "09", "12")

enr_grades <- fetch_enr_multi(2019:2025) |>
  filter(
    is_state,
    subgroup == "total_enrollment",
    grade_level %in% grade_levels
  ) |>
  mutate(
    grade_label = case_when(
      grade_level == "K" ~ "Kindergarten",
      grade_level == "01" ~ "1st Grade",
      grade_level == "05" ~ "5th Grade",
      grade_level == "09" ~ "9th Grade",
      grade_level == "12" ~ "12th Grade"
    ),
    grade_label = factor(grade_label,
      levels = c("Kindergarten", "1st Grade", "5th Grade", "9th Grade", "12th Grade"))
  ) |>
  group_by(grade_label) |>
  mutate(
    baseline = first(n_students),
    index = n_students / baseline * 100
  ) |>
  ungroup()

ggplot(enr_grades, aes(x = end_year, y = index, color = grade_label)) +
  geom_hline(yintercept = 100, linetype = "dashed", color = "gray50") +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = 2019:2025) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Grade Level Enrollment Trends",
    subtitle = "Index: 2019 = 100",
    x = "School Year (End)",
    y = "Index (2019 = 100)",
    color = "Grade"
  ) +
  theme(legend.position = "bottom")
```

## 6. District Size Distribution

Oklahoma has many small rural districts alongside large urban systems.

```{r district-size-chart}
# District size distribution
enr_size <- fetch_enr(2025) |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  mutate(
    size_bucket = case_when(
      n_students < 100 ~ "Under 100",
      n_students < 500 ~ "100-499",
      n_students < 1000 ~ "500-999",
      n_students < 5000 ~ "1,000-4,999",
      n_students < 10000 ~ "5,000-9,999",
      TRUE ~ "10,000+"
    ),
    size_bucket = factor(size_bucket,
      levels = c("Under 100", "100-499", "500-999", "1,000-4,999", "5,000-9,999", "10,000+"))
  ) |>
  count(size_bucket)

ggplot(enr_size, aes(x = size_bucket, y = n, fill = size_bucket)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = n), vjust = -0.5, size = 4) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  scale_fill_brewer(palette = "Blues") +
  labs(
    title = "Oklahoma District Size Distribution",
    subtitle = "Number of districts by enrollment size (2024-25)",
    x = "District Size (Students)",
    y = "Number of Districts"
  )
```

## 7. Tulsa Metro Area Trends

The Tulsa metro area includes Union, Jenks, and Broken Arrow - fast-growing suburban districts.

```{r tulsa-metro-chart}
# Tulsa area districts comparison
tulsa_districts <- c("72I001", "72I009", "72I005", "09I001")
tulsa_names <- c(
  "72I001" = "Tulsa",
  "72I009" = "Union",
  "72I005" = "Jenks",
  "09I001" = "Broken Arrow"
)

enr_tulsa <- fetch_enr_multi(2016:2025) |>
  filter(
    district_id %in% tulsa_districts,
    is_district,
    subgroup == "total_enrollment",
    grade_level == "TOTAL"
  ) |>
  mutate(district_label = tulsa_names[district_id])

ggplot(enr_tulsa, aes(x = end_year, y = n_students, color = district_label)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(breaks = seq(2016, 2025, 2)) +
  scale_color_manual(values = c(
    "Tulsa" = "#841617",
    "Union" = "#1a5276",
    "Jenks" = "#148f77",
    "Broken Arrow" = "#9c640c"
  )) +
  labs(
    title = "Tulsa Metro Area Districts",
    subtitle = "Enrollment trends 2016-2025",
    x = "School Year (End)",
    y = "Students",
    color = "District"
  ) +
  theme(legend.position = "bottom")
```

## 8. County Enrollment Concentration

Enrollment is heavily concentrated in a few counties, with Oklahoma and Tulsa counties accounting for nearly a third of all students.

```{r county-concentration-chart}
# Top 10 counties by enrollment
enr_county <- fetch_enr(2025) |>
  filter(is_district, grade_level == "TOTAL", subgroup == "total_enrollment") |>
  group_by(county) |>
  summarize(n_students = sum(n_students, na.rm = TRUE), .groups = "drop") |>
  filter(!is.na(county)) |>
  arrange(desc(n_students)) |>
  head(10)

ggplot(enr_county, aes(x = reorder(county, n_students), y = n_students)) +
  geom_col(fill = "#841617") +
  geom_text(aes(label = scales::comma(n_students)), hjust = -0.1, size = 3.5) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.15))) +
  coord_flip() +
  labs(
    title = "Top 10 Oklahoma Counties by Enrollment",
    subtitle = "2024-25 school year",
    x = NULL,
    y = "Total Students"
  )
```

## 9. Limited English Proficiency Trends

Oklahoma has seen steady growth in Limited English Proficiency (LEP) enrollment over the past decade, reflecting demographic changes across the state.

```{r lep-trend-chart}
lep_trend <- fetch_enr_multi(2016:2025) |>
  filter(is_state, subgroup == "lep") |>
  mutate(n_students = sum(n_students, na.rm = TRUE)) |>
  select(end_year, n_students) |>
  distinct()

ggplot(lep_trend, aes(x = end_year, y = n_students)) +
  geom_line(color = "#1a5276", linewidth = 1) +
  geom_point(color = "#1a5276", size = 3) +
  scale_x_continuous(breaks = 2016:2025) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Statewide LEP Enrollment Growth",
    subtitle = "Total students with Limited English Proficiency (2015-16 to 2024-25)",
    x = "School Year (End Year)",
    y = "LEP Students"
  )
```

## 10. Rural Consolidation Patterns

The smallest districts are concentrated in rural western and southeastern Oklahoma.

```{r small-districts-chart}
# Enrollment trends in small rural districts vs. state
small_districts <- fetch_enr(2025) |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  filter(n_students < 200) |>
  pull(district_id)

enr_small <- fetch_enr_multi(2016:2025) |>
  filter(
    is_district | is_state,
    subgroup == "total_enrollment",
    grade_level == "TOTAL"
  ) |>
  mutate(category = case_when(
    is_state ~ "State Total",
    district_id %in% small_districts ~ "Small Districts (<200)",
    TRUE ~ "Other Districts"
  )) |>
  group_by(end_year, category) |>
  summarize(n_students = sum(n_students), .groups = "drop") |>
  group_by(category) |>
  mutate(index = n_students / first(n_students) * 100) |>
  ungroup()

ggplot(enr_small |> filter(category != "Other Districts"),
       aes(x = end_year, y = index, color = category)) +
  geom_hline(yintercept = 100, linetype = "dashed", color = "gray50") +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = seq(2016, 2025, 2)) +
  scale_color_manual(values = c("State Total" = "#1a5276", "Small Districts (<200)" = "#841617")) +
  labs(
    title = "Small District Enrollment Decline",
    subtitle = "Index: 2016 = 100",
    x = "School Year (End)",
    y = "Index (2016 = 100)",
    color = NULL
  ) +
  theme(legend.position = "bottom")
```

## 11. EPIC Charter Schools: Virtual Learning Giant

EPIC Charter Schools has become one of Oklahoma's largest educational entities, growing rapidly through virtual learning.

```{r epic-charter-chart}
# Find EPIC districts
epic_ids <- fetch_enr(2025) |>
  filter(is_district, grepl("EPIC", district_name, ignore.case = TRUE)) |>
  distinct(district_id) |>
  pull(district_id)

# If EPIC found, show trend
if (length(epic_ids) > 0) {
  enr_epic <- fetch_enr_multi(2016:2025) |>
    filter(
      district_id %in% epic_ids,
      is_district,
      subgroup == "total_enrollment",
      grade_level == "TOTAL"
    ) |>
    group_by(end_year) |>
    summarize(n_students = sum(n_students), .groups = "drop")

  ggplot(enr_epic, aes(x = end_year, y = n_students)) +
    geom_area(fill = "#841617", alpha = 0.3) +
    geom_line(color = "#841617", linewidth = 1.2) +
    geom_point(color = "#841617", size = 3) +
    scale_y_continuous(labels = scales::comma, limits = c(0, NA)) +
    scale_x_continuous(breaks = 2016:2025) +
    labs(
      title = "EPIC Charter Schools Enrollment",
      subtitle = "Virtual charter growth in Oklahoma",
      x = "School Year (End)",
      y = "Students"
    )
} else {
  # Fallback: Show all charter growth
  charter_enr <- fetch_enr_multi(2016:2025) |>
    filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") |>
    filter(grepl("Charter|Virtual|Academy", district_name, ignore.case = TRUE)) |>
    group_by(end_year) |>
    summarize(n_students = sum(n_students), .groups = "drop")

  ggplot(charter_enr, aes(x = end_year, y = n_students)) +
    geom_area(fill = "#841617", alpha = 0.3) +
    geom_line(color = "#841617", linewidth = 1.2) +
    geom_point(color = "#841617", size = 3) +
    scale_y_continuous(labels = scales::comma, limits = c(0, NA)) +
    scale_x_continuous(breaks = 2016:2025) +
    labs(
      title = "Charter and Virtual School Enrollment",
      subtitle = "Alternative school sector growth in Oklahoma",
      x = "School Year (End)",
      y = "Students"
    )
}
```

## 12. Black Student Enrollment Distribution

Black students are concentrated in urban districts, particularly in Oklahoma City and Tulsa.

```{r black-enrollment-chart, eval=FALSE}
# Districts with highest black enrollment
# NOTE: Race data not available in OSDE enrollment files
# This chart is disabled pending data source updates
enr_black <- fetch_enr(2025) |>
  filter(is_district, grade_level == "TOTAL",
         subgroup %in% c("total_enrollment", "black")) |>
  select(district_id, district_name, subgroup, n_students) |>
  pivot_wider(names_from = subgroup, values_from = n_students,
              values_fn = sum) |>
  filter(!is.na(total_enrollment), total_enrollment >= 500,
         !is.na(black), black > 0) |>
  mutate(pct_black = black / total_enrollment) |>
  arrange(desc(pct_black)) |>
  head(12)

ggplot(enr_black, aes(x = reorder(district_name, pct_black), y = pct_black)) +
  geom_col(fill = "#148f77") +
  geom_text(aes(label = scales::percent(pct_black, accuracy = 1)), hjust = -0.1, size = 3.5) +
  scale_y_continuous(labels = scales::percent, expand = expansion(mult = c(0, 0.15))) +
  coord_flip() +
  labs(
    title = "Districts with Highest Black Enrollment",
    subtitle = "Percentage of total (2024-25, districts with 500+ students)",
    x = NULL,
    y = "Percent Black"
  )
```

## 13. Southeast Oklahoma: Poverty Corridor

The southeastern region shows high economic disadvantage rates and declining enrollment.

```{r southeast-chart}
# Counties in southeast Oklahoma
se_counties <- c("McCurtain", "Pushmataha", "Choctaw", "LeFlore", "Latimer",
                 "Pittsburg", "Atoka", "Bryan", "Coal", "Haskell")

# Get total enrollment by county
enr_se <- fetch_enr(2025) |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  mutate(region = if_else(county %in% se_counties, "Southeast", "Rest of State")) |>
  group_by(region) |>
  summarize(
    districts = n(),
    students = sum(n_students),
    .groups = "drop"
  ) |>
  mutate(pct = students / sum(students))

# Show regional comparison
enr_region <- fetch_enr_multi(2016:2025) |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  mutate(region = if_else(county %in% se_counties, "Southeast", "Rest of State")) |>
  group_by(end_year, region) |>
  summarize(n_students = sum(n_students), .groups = "drop") |>
  group_by(region) |>
  mutate(index = n_students / first(n_students) * 100) |>
  ungroup()

ggplot(enr_region, aes(x = end_year, y = index, color = region)) +
  geom_hline(yintercept = 100, linetype = "dashed", color = "gray50") +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = seq(2016, 2025, 2)) +
  scale_color_manual(values = c("Southeast" = "#841617", "Rest of State" = "#1a5276")) +
  labs(
    title = "Southeast Oklahoma vs. Rest of State",
    subtitle = "Enrollment index: 2016 = 100",
    x = "School Year (End)",
    y = "Index (2016 = 100)",
    color = NULL
  ) +
  theme(legend.position = "bottom")
```

## 14. Kindergarten Enrollment Over Time

Kindergarten enrollment is a leading indicator of future cohort sizes moving through the system.

```{r kindergarten-trend-chart}
# Kindergarten enrollment trend
enr_k <- fetch_enr_multi(2016:2025) |>
  filter(is_state, subgroup == "total_enrollment", grade_level == "K")

ggplot(enr_k, aes(x = end_year, y = n_students)) +
  geom_line(color = "#841617", linewidth = 1.2) +
  geom_point(color = "#841617", size = 3) +
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(breaks = 2016:2025) +
  labs(
    title = "Oklahoma Kindergarten Enrollment",
    subtitle = "Statewide kindergarten class size 2016-2025",
    x = "School Year (End)",
    y = "Kindergarten Students"
  )
```

## 15. The Panhandle: Oklahoma's Remote Northwest

The three panhandle counties (Cimarron, Texas, Beaver) have unique enrollment patterns shaped by agriculture and isolation.

```{r panhandle-chart}
# Panhandle counties
panhandle_counties <- c("Cimarron", "Texas", "Beaver")

# Compare panhandle to state trends
enr_panhandle <- fetch_enr_multi(2016:2025) |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  mutate(region = case_when(
    county %in% panhandle_counties ~ "Panhandle",
    TRUE ~ "Rest of State"
  )) |>
  group_by(end_year, region) |>
  summarize(n_students = sum(n_students), .groups = "drop") |>
  group_by(region) |>
  mutate(index = n_students / first(n_students) * 100) |>
  ungroup()

ggplot(enr_panhandle, aes(x = end_year, y = index, color = region)) +
  geom_hline(yintercept = 100, linetype = "dashed", color = "gray50") +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = seq(2016, 2025, 2)) +
  scale_color_manual(values = c("Panhandle" = "#9c640c", "Rest of State" = "#1a5276")) +
  labs(
    title = "Oklahoma Panhandle Enrollment Trends",
    subtitle = "Three-county region (Cimarron, Texas, Beaver) vs. state",
    x = "School Year (End)",
    y = "Index (2016 = 100)",
    color = NULL
  ) +
  theme(legend.position = "bottom")
```

## Summary

Oklahoma's school enrollment data reveals several key trends:

1. **Post-COVID recovery**: Statewide enrollment has rebounded from the 2021 dip
2. **Urban-suburban shift**: Oklahoma City is losing students while Edmond and Moore grow
3. **Many small districts**: Over 200 districts have fewer than 500 students
4. **Tulsa metro growth**: Suburban districts like Union, Jenks, and Broken Arrow are growing
5. **County concentration**: Oklahoma and Tulsa counties dominate statewide enrollment
6. **Virtual learning explosion**: EPIC Charter Schools grew into one of the state's largest districts
7. **Regional disparities**: Southeast Oklahoma faces population decline while metro areas grow
8. **Panhandle isolation**: Remote northwestern counties show distinct patterns
9. **Kindergarten as bellwether**: Early grade enrollment trends predict future system capacity needs
10. **English learner concentration**: LEP students are concentrated in specific urban districts

Use `fetch_enr()` and `fetch_enr_multi()` to explore these patterns further in your own analysis.
